### 2010-09-28 Thesis/Chapter5.tex
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 28 September -
Major update: lot of fixes on the first part that has been reviewed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@177 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

	\subsection{Computation of node displacements} \label{chap5:computeNodalDisp}
We recall that the dynamic system of equations describing a deformable solid is the following (see section \ref{chap3:dynamic2}:
\begin{equation}
\label{chap5:eqDynamic}
\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
\end{equation}		
where $ \mathbf{M} $ is a constant mass matrix, $\mathbf{D}$ is a constant damping matrix, $ \mathbf{K}^e(\mathbf{U}^e) $ is the stiffness matrix, which is a function of nodal displacements $\mathbf{U}$, and $\mathbf{R}$ are externally applied loads. And we seek the unknown displacements $ \mathbf{U} $.

@@

Of crucial importance in explicit analyses is the restriction on time step sizes imposed by stability constraints \label{chap5:discussionTimestep}. As mentioned, the central difference method is only stable if $ \Delta t <  \Delta t_{cr} $. Essentially, stability requires that the time step be not larger than the time required for a dilatational wave to propagate through the smallest element. In linear analyses this translates into the following formula for $ \Delta t_{cr}$:
\begin{equation}
\label{chap5:criticalTimestep}
\Delta t_{cr} = \dfrac{L_e}{c},
\end{equation}
where $ L_e $ is the smallest characteristic element length (roughly interpreted as the smallest edge length) in the assembly, and $c$ is the dilatational wave speed of the material, given by
\begin{equation}
\label{chap5:waveSpeed}
c = \sqrt{\dfrac{E (1 - \nu)}{\rho (1+\nu)(1-2\nu)}},
\end{equation}
where $E$ is Young's modulus and $\nu$ is Poisson's ratio. If nonlinear analyses (either geometric or material) are conducted, a somewhat smaller time step must be employed since $c$ will change with deformation, and in particular will generally increase. Additionally, the presence of damping necessitates a further decrease in the limit. Nonetheless, \eqref{chap5:waveSpeed} may be used to estimate $ \Delta t_{cr}$.

Some important observations arise from \eqref{chap5:criticalTimestep} and \eqref{chap5:waveSpeed}. Firstly, since the stiffness of soft tissues is very much smaller than that of common engineering materials ($E \approx 3 \times 10^3\, $Pa for brain tissue versus $\approx 2 \times 10^{11}\, $Pa for steel), the allowable time step is much larger for analysis of the former. This is a key reason for the expediency of explicit time integration for analysis of soft tissues. Secondly, soft tissues are generally considered to be incompressible, leading to $\nu \approx 0.49$ commonly being employed. However, if this value can be relaxed even further, significant increases in $ \Delta t_{cr}$ are possible. As an example, for a given $ Le $ a material with $E = 3\,000\, $Pa and $\nu = 0.45$ allows a time step of more than double that of a material with $\nu = 0.49$. Of course, lowering the Poisson ratio introduces inaccuracies, but for some applications (interactive medical simulation for instance) it may be that this is acceptable. Finally, it must be highlighted that $ \Delta t_{cr}$ is proportional to $ L_e $. If the element size is decreased (by using a finer mesh, or by analysing very small objects), inexorably $ \Delta t_{cr}$ is also reduced.

>>>>

	\subsection{Computation of node displacements} \label{chap5:computeNodalDisp}
We recall that the dynamic system of equations describing a deformable solid is the following (see section \ref{chap3:dynamic2}):
\begin{equation}
\label{chap5:eqDynamic}
\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
\end{equation}		
where $ \mathbf{M} $ is a constant mass matrix, $\mathbf{D}$ is a constant damping matrix, $ \mathbf{K}^e(\mathbf{U}^e) $ is the stiffness matrix, which is a function of nodal displacements $\mathbf{U}$, and $\mathbf{R}$ are externally applied loads. And we seek the unknown displacements $ \mathbf{U} $.

@@

Of crucial importance in explicit analyses is the restriction on time step sizes imposed by stability constraints\label{chap5:discussionTimestep}. As mentioned, the central difference method is only stable if $ \Delta t <  \Delta t_{cr} $. Essentially, stability requires that the time step be not larger than the time required for a dilatational wave to propagate through the smallest element. In linear analyses this translates into the following formula for $ \Delta t_{cr}$:
\begin{equation}
\label{chap5:criticalTimestep}
\Delta t_{cr} = \dfrac{L_e}{c},
\end{equation}
where $ L_e $ is the smallest characteristic element length (roughly interpreted as the smallest edge length) in the assembly, and $c$ is the dilatational wave speed of the material, given by
\begin{equation}
\label{chap5:waveSpeed}
c = \sqrt{\dfrac{E (1 - \nu)}{\rho (1+\nu)(1-2\nu)}},
\end{equation}
where $E$ is Young's modulus and $\nu$ is Poisson's ratio. If nonlinear analyses (either geometric or material) are conducted, a somewhat smaller time step must be employed since $c$ will change with deformation, and in particular will generally increase. Additionally, the presence of damping necessitates a further decrease in the limit. Nonetheless, \eqref{chap5:waveSpeed} may be used to estimate $ \Delta t_{cr}$.

Some important observations arise from \eqref{chap5:criticalTimestep} and \eqref{chap5:waveSpeed}. Firstly, since the stiffness of soft tissues is very much smaller than that of common engineering materials ($E \approx 3 \times 10^3\, $Pa for brain tissue versus $\approx 2 \times 10^{11}\, $Pa for steel), the allowable time step is much larger for analysis of the former. This is a key reason for the expediency of explicit time integration for analysis of soft tissues. Secondly, soft tissues are generally considered to be incompressible, leading to $\nu \approx 0.49$ commonly being employed. However, if this value can be relaxed even further, significant increases in $ \Delta t_{cr}$ are possible. As an example, for a given $ L_e $ a material with $E = 3\,000\, $Pa and $\nu = 0.45$ allows a time step of more than double that of a material with $\nu = 0.49$. Of course, lowering the Poisson ratio introduces inaccuracies, but for some applications (interactive medical simulation for instance) it may be that this is acceptable. Finally, it must be highlighted that $ \Delta t_{cr}$ is proportional to $ L_e $. If the element size is decreased (by using a finer mesh, or by analysing very small objects), inexorably $ \Delta t_{cr}$ is also reduced.

### 2010-09-28 Thesis/Chapter6.tex
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 28 September -
Major update: lot of fixes on the first part that has been reviewed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@177 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\section{General-purpose computation on graphics processing units}

@@

In the precomputation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel. We add nodes to a block along with their unique elements until the shared memory is saturated. Of course, in this approach some elements will be computed several times (because being into different blocks). However, storing the force contributions in shared memory instead of global memory to read them later on via texture fetches widely outperforms the need of computing several elements several times.

@@

The first factor leads to a strong constraint on the quality of the mesh. Indeed, meshing a complex shape with good and uniform elements is not a straightforward task and rather often bad and very small elements are created. In such a case, the critical time step for the whole simulation would be small even if there is only a single small element in the mesh. As an example, the liver mesh used in the section~\ref{chap6:liverSimu} had to be smoothed and remeshed properly to allow a decent simulation.

@@

	\subsection{Handling contacts}
The present work was entirely concerned with numerical solution of constitutive models in explicit FE analyses, however an equally important issue from the point of view of interactive simulation is the modelling of interaction between organs and virtual tools, and its possible impact on $ \Delta t_{cr} $ and the stability of the simulation. Explicit time integration schemes are known to be more sensitive to contacts than implicit schemes. While this is beyond the scope of the present work, we note that a range of contact formulations are available in industry standard commercial explicit FE codes (e.g. LS-Dyna and ABAQUS). The methods are well tested and understood in the computational mechanics community, with two main types in common use: penalty methods and kinematic constraint methods. Penalty methods involve use of fictitious contact springs at interfaces, effectively to penalise interpenetration of opposing surfaces. If excessively stiff springs are used $ \Delta t_{cr} $ may indeed be reduced, however except in extreme conditions (e.g. from explosive or high speed impacts, or imposition of very large forces on highly constrained objects) spring stiffnesses of the same order of magnitude as the material stiffness are suitable. The effect on $ \Delta t_{cr} $ is therefore minimal \citep{Hallquist06}. Kinematic constraint methods involve (loosely) direct relocation of penetrating nodes back to the outside surface of the penetrated body, and have no effect on $ \Delta t_{cr} $. These may be especially useful for modelling interaction between soft tissues and stiff (effectively rigid) surgical tools. The computational cost of either variety has not been investigated here.

>>>>

\section{General-purpose computation on GPU}

@@

In the precomputation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel. We add nodes to a block along with their unique elements until the shared memory is saturated. Of course, in this approach some elements will be computed several times (because being into different blocks). \ON Yet, \OFF storing the force contributions in shared memory instead of global memory to read them later on via texture fetches \ON may \OFF outperform the need of computing several elements several times. \ON However, the efficiency of this single kernel approach depends on the nature of the mesh. Because the force contributions of all elements attached to a node must be eventually added up, this approach allows to avoid the numerous (slow) accesses to global memory when the valency of the nodes is high (that is, nodes are attached to many elements). This type of mesh is often encountered when dealing with the complex shapes of anatomical structures. Therefore, where the first implementation showed its limits with complex meshes, this single kernel approach has proven to be more computationally effective. 
\OFF

@@

The first factor leads to a strong constraint on the quality of the mesh. Indeed, meshing a complex shape with good and uniform elements is not a straightforward task and rather often bad and very small elements are created. In such a case, the critical time step for the whole simulation would be small even if there is only a single small element in the mesh. As an example, the liver mesh used in the section~\ref{chap6:liverSimu} had to be smoothed and remeshed properly \ON with great care \OFF to allow a decent simulation.

@@

	\subsection{Handling contacts}
The present work was entirely concerned with numerical solution of constitutive models in explicit FE analyses, however an equally important issue from the point of view of interactive simulation is the modelling of interaction between organs and virtual tools, and its possible impact on $ \Delta t_{cr} $ and the stability of the simulation. Explicit time integration schemes are known to be more sensitive to contacts than implicit schemes. While this is beyond the scope of the present work, we note that a range of contact formulations are available, with two main types in common use: penalty methods and kinematic constraint methods. Penalty methods involve use of fictitious contact springs at interfaces, effectively to penalise interpenetration of opposing surfaces. If excessively stiff springs are used, $ \Delta t_{cr} $ may indeed be reduced. \ON According to LS-Dyna documentation \citep{Hallquist06},  \OFF except in extreme conditions (that is from explosive or high speed impacts, or imposition of very large forces on highly constrained objects), spring stiffnesses of the same order of magnitude as the material stiffness are suitable and the effect on $ \Delta t_{cr} $ is therefore minimal. \ON However, in practice, we observed that spring stiffnesses greater than the material stiffness are often required to prevent any penetration of the two objects, which results in an unstable system. \OFF Kinematic constraint methods involve (loosely) direct relocation of penetrating nodes back to the outside surface of the penetrated body, and have no effect on $ \Delta t_{cr} $. \ON In theory, \OFF these may be especially useful for modelling interaction between soft tissues and stiff (effectively rigid) surgical tools. \ON It is worth nothing that \OFF the computational cost of either variety has not been investigated here.

\ON
Even if a stable contact modelling could be achieved, one element of crucial importance has been omitted so far: the collision detection process. Indeed, before even handling the interaction between two objects, their interaction needs to be detected. Because the time step must be very small when using an explicit time integration scheme (often around one millisecond or less), the collision detection algorithm must also be run at the same frequency. The problem is that no collision detection algorithm is fast enough to execute at $ 1\,000\,$Hz or more with objects of reasonable sizes. Of course, one could choose to detect the collisions at a lesser frequency but the risks of penetration would substantially increases. One notable exception is the interaction between a soft deformable object and a fixed rigid surface where the collision detection is rapid. 
\OFF

\TODO{timestep monitoring}

\ON
	\subsubsection{Conclusion}
As we showed in the last two chapters, the TLED algorithm is a very efficient fully non-linear finite element formulation. We also added viscoelasticity and anisotropy features with minimal impact on its performance. In particular, its GPU implementation allows non-linear analyses of structures in real-time using a high number of elements as never before.

However, the use of an explicit integration scheme imposes substantial constraints which limits TLED's range of applications. More specifically, the usage of the TLED algorithm in very interactive environments (as in the manipulation of a liver with a tool) is not recommended. The collision detection would either become the bottleneck of the simulation (if run at every time step) or fail to provide accurate detection of contacts (otherwise). In addition, the TLED algorithm is also not appropriate to simulate the deformation of rigid object as the critical time step would be too small to allow real-time computations. In constrast, TLED excels for computing the deformation of soft tissues when the boundary conditions are controlled. As an exemple, the TLED algorithm may efficiently provide brain shift predictions as the interaction of the brain with the fixed skull is determined beforehand and rapid collision detection can be achieved in this case.

Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs.
\OFF
